---
title: "Cellular Automata Tutorial"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Cellular Automata Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(emergenceModelR)
```

## Purpose

This tutorial introduces `simulate_cellular_automata()`. Cellular automata are useful because they show how simple local rules can produce unexpectedly rich global patterns.

In this tutorial, you will learn how to:

- run a basic cellular automaton;
- inspect the output;
- visualize cellular automata patterns;
- compare Rule 30 and Rule 110;
- summarize outputs using `measure_emergence()`;
- interpret cellular automata as simplified models of emergence.

## What the function represents

A cellular automaton is a grid of cells that update over time according to a rule.

In an elementary cellular automaton:

- cells are arranged in one dimension;
- each cell has a binary state, usually `0` or `1`;
- each cell updates based on itself and its nearby neighbors;
- the same rule is applied repeatedly across the system.

The function `simulate_cellular_automata()` creates this kind of simplified model.

The main idea is:

> Local update rules can generate global space-time patterns.

## Main arguments

| Argument | Meaning |
|---|---|
| `rule` | The cellular automaton rule number |
| `n_cells` | Number of cells in the row |
| `steps` | Number of time steps to simulate |

The `rule` argument is especially important. Different rules can produce very different global patterns, even when the number of cells and steps stays the same.

## Basic simulation: Rule 30

Rule 30 is often used to show how a simple deterministic rule can produce irregular-looking patterns.

```{r ca-rule30}
rule30 <- simulate_cellular_automata(
  rule = 30,
  n_cells = 81,
  steps = 80
)

head(rule30)
```

## Inspect the output

The output is a data frame.

```{r ca-rule30-structure}
str(rule30)
```

The main columns usually include:

| Column | Meaning |
|---|---|
| `step` | Time step |
| `cell` | Cell position |
| `state` | Cell state, usually `0` or `1` |

This structure makes it easy to visualize the pattern across cells and time.

## Plot Rule 30

```{r ca-rule30-plot}
plot_emergence_sim(
  rule30,
  x = "cell",
  y = "step",
  value = "state",
  type = "raster"
)
```

## Interpretation of Rule 30

The pattern is generated by repeatedly applying a local rule. No cell knows the global pattern. No central controller designs the structure.

This illustrates a core idea of emergence:

> A simple local rule can generate a complex system-level pattern.

Rule 30 is useful for teaching because the output can look surprisingly irregular even though the rule is deterministic.

## Rule 110

Rule 110 is another important cellular automaton rule. It is often discussed because it can produce structured patterns associated with computation-like behavior.

```{r ca-rule110}
rule110 <- simulate_cellular_automata(
  rule = 110,
  n_cells = 81,
  steps = 80
)

head(rule110)
```

```{r ca-rule110-plot}
plot_emergence_sim(
  rule110,
  x = "cell",
  y = "step",
  value = "state",
  type = "raster"
)
```

## Interpretation of Rule 110

Rule 110 often produces more visibly structured patterns than Rule 30. These structures may persist, move, or interact across time.

The important teaching point is not that Rule 110 is alive or conscious. Rather, it shows that very simple rule-based systems can generate patterns with rich dynamics.

## Compare metrics

The function `measure_emergence()` provides simple emergence-oriented summaries.

```{r ca-metrics}
rbind(
  rule30 = measure_emergence(
    rule30,
    value_col = "state",
    time_col = "step"
  ),
  rule110 = measure_emergence(
    rule110,
    value_col = "state",
    time_col = "step"
  )
)
```

## Interpreting the metrics

The metrics summarize selected features of the cellular automata outputs, such as diversity, variability, or temporal change.

They are useful for comparison, but they should not be treated as a complete measure of emergence.

A careful interpretation is:

> The metrics help compare patterns produced by different rules.

An overstatement would be:

> The metrics prove that one rule is truly more emergent.

The first statement is appropriate. The second is too strong.

## Compare another rule

Try Rule 90, which often produces a more regular nested pattern.

```{r ca-rule90}
rule90 <- simulate_cellular_automata(
  rule = 90,
  n_cells = 81,
  steps = 80
)

plot_emergence_sim(
  rule90,
  x = "cell",
  y = "step",
  value = "state",
  type = "raster"
)
```

## Compare three rules

```{r ca-three-rule-metrics}
rbind(
  rule30 = measure_emergence(
    rule30,
    value_col = "state",
    time_col = "step"
  ),
  rule90 = measure_emergence(
    rule90,
    value_col = "state",
    time_col = "step"
  ),
  rule110 = measure_emergence(
    rule110,
    value_col = "state",
    time_col = "step"
  )
)
```

## Interpretation of rule comparison

All three simulations use the same number of cells and the same number of steps. The main difference is the local update rule.

This makes cellular automata useful for teaching parameter sensitivity:

> Changing the rule can change the global pattern.

The components are simple, but the system-level behavior can vary greatly.

## Change the number of cells

The argument `n_cells` controls the width of the cellular automaton.

```{r ca-change-cells}
rule30_wide <- simulate_cellular_automata(
  rule = 30,
  n_cells = 121,
  steps = 80
)

plot_emergence_sim(
  rule30_wide,
  x = "cell",
  y = "step",
  value = "state",
  type = "raster"
)
```

## Interpretation of `n_cells`

Increasing `n_cells` gives the pattern more space to develop horizontally. This can make global structure easier to see.

However, the local rule is unchanged. The same rule is being applied across a larger system.

## Change the number of steps

The argument `steps` controls how long the system evolves.

```{r ca-change-steps}
rule30_long <- simulate_cellular_automata(
  rule = 30,
  n_cells = 81,
  steps = 120
)

plot_emergence_sim(
  rule30_long,
  x = "cell",
  y = "step",
  value = "state",
  type = "raster"
)
```

## Interpretation of `steps`

Increasing `steps` allows the pattern to unfold for a longer period. This is important because emergent patterns often become visible only through repeated updating over time.

The system-level structure is not present all at once. It develops through iteration.

## Suggested exercises

Try modifying the examples above.

```{r ca-exercise}
experiment <- simulate_cellular_automata(
  rule = 150,
  n_cells = 101,
  steps = 100
)

plot_emergence_sim(
  experiment,
  x = "cell",
  y = "step",
  value = "state",
  type = "raster"
)
```

Questions to consider:

- How does Rule 30 differ from Rule 110?
- Which rules produce regular-looking patterns?
- Which rules produce irregular-looking patterns?
- What happens when you increase `n_cells`?
- What happens when you increase `steps`?
- Do the metrics match what you see in the plots?

## Responsible interpretation

Cellular automata are simplified educational models. They are useful because they make the relationship between local rules and global patterns easy to observe.

It is better to say:

> The simulation illustrates how local update rules can generate emergent patterns.

than:

> The simulation fully explains emergence in real systems.

Cellular automata are powerful teaching tools, but they are highly abstract.

## Key takeaway

`simulate_cellular_automata()` helps users explore how simple local rules can generate complex global patterns.

Rule 30 shows how deterministic rules can produce irregular-looking structure. Rule 110 shows how simple rules can produce richer, computation-like patterns. Together, they demonstrate one of the central lessons of emergence:

> Global structure can arise from repeated local updating without central control.

