---
title: "Self-Organization"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Self-Organization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load-package}
library(emergenceModelR)
```

## Purpose

This article explains self-organization as a major mechanism of emergence. Self-organization occurs when order, structure, or pattern develops through internal system dynamics rather than through external design or centralized control [@prigogine1984order; @kauffman1993origins; @mitchell2009complexity].

The purpose of this chapter is to show how local interactions, feedback, diffusion, and repeated updating can generate large-scale organization. In `emergenceModelR`, this idea is represented by `simulate_self_organization()`, a simplified grid model that allows learners to observe how spatial structure can arise from distributed dynamics.

The guiding question is:

> How can order arise in a system without a central organizer?

## What is self-organization?

Self-organization refers to the spontaneous formation of pattern or structure through the interactions of a system’s own components. The word “spontaneous” does not mean random or unexplained. It means that the organization is not imposed from outside by a central controller.

In a self-organizing system, the global pattern arises because many local units influence one another over time. No single component contains the complete pattern. No part of the system needs a blueprint of the final form. Instead, the pattern emerges through repeated interaction.

Examples often discussed in complexity science include:

- chemical pattern formation;
- convection cells;
- biological morphogenesis;
- flocking and swarming;
- ant-colony organization;
- ecosystem dynamics;
- market coordination;
- neural synchronization.

These examples differ greatly in their physical details, but they share a general logic: local interactions produce system-level order.

## Self-organization and emergence

Self-organization and emergence are closely related, but they are not identical.

**Emergence** refers to the appearance of system-level patterns or properties arising from lower-level interactions.

**Self-organization** refers to one process by which such patterns arise: the system organizes itself through internal dynamics.

In this sense, self-organization can be understood as a mechanism of emergence. Emergence describes the relationship between levels. Self-organization describes how order can form without centralized control.

| Concept | Main emphasis | Example question |
|---|---|---|
| Emergence | System-level patterns arising from local interactions | What global pattern appears? |
| Self-organization | Formation of order through internal dynamics | How does the system organize itself? |
| Complexity | Many interacting parts with nonlinear dynamics | Why is the system difficult to predict? |

`simulate_self_organization()` focuses on the second question: how distributed local updating can produce spatial order.

## Conceptual background

Self-organizing systems are often composed of many interacting parts. The parts may be molecules, cells, organisms, agents, nodes, or spatial locations in a grid. Each part follows relatively simple rules, but the repeated application of those rules can produce large-scale organization.

Three concepts are especially important:

1. **Local interaction**: each part responds to nearby conditions rather than to the whole system.
2. **Feedback**: current states influence future states, sometimes amplifying differences.
3. **Constraint**: system structure, boundaries, and environmental conditions limit what patterns can form.

The key point is that order can arise from dynamics. The system does not need to be externally arranged into the final pattern. Instead, structure develops through the way the system changes over time.

## Feedback and diffusion

Two important mechanisms in many self-organizing systems are feedback and diffusion.

**Feedback** occurs when the current state of a system influences its future state. Positive feedback amplifies local differences. Negative feedback stabilizes or limits change.

**Diffusion** spreads or smooths local variation. In physical and biological systems, diffusion may involve the movement of heat, chemicals, organisms, information, or influence across space.

Complex patterns often arise from the tension between amplification and smoothing. Feedback can intensify local structure. Diffusion can spread that structure outward or reduce sharp differences.

This interaction is central to many pattern-forming systems. If feedback dominates completely, the system may become unstable or overly concentrated. If diffusion dominates completely, the system may become too uniform. Interesting patterns often appear between these extremes.

## Non-equilibrium order

Self-organization is often associated with systems far from equilibrium. In equilibrium, differences tend to disappear and the system becomes stable or uniform. Far from equilibrium, flows of energy, matter, or information can maintain organized structure [@prigogine1984order].

This is important for life and complexity. Living systems are not passive equilibrium structures. They maintain organization through continuous flows: energy, nutrients, waste, signals, and regulation. Self-organization provides one way to think about how order can be maintained in dynamic systems.

The simplified model in this package does not simulate real thermodynamics. However, it illustrates the general idea that repeated updating and interaction can generate structured outcomes.

## Relation to the package

`simulate_self_organization()` represents self-organization in a simplified spatial grid. Each grid cell has a numeric value. Over time, values change through local updating influenced by diffusion and feedback.

| Theoretical concept | Package representation |
|---|---|
| Spatial system | Grid |
| Local state | Cell value |
| Local interaction | Updating based on nearby values |
| Diffusion | Smoothing/spreading parameter |
| Feedback | Amplification/reinforcement parameter |
| System evolution | Repeated time steps |
| Emergent pattern | Final spatial distribution |

The model is not intended to represent a specific chemical or biological system. It is an educational abstraction for exploring how local updating can generate global pattern.

## Simulation example

```{r theory-so-example}
so <- simulate_self_organization(
  grid_size = 30,
  steps = 50,
  diffusion = 0.20,
  feedback = 0.60,
  seed = 2
)

final_step <- subset(so, step == max(step))

plot_emergence_sim(
  final_step,
  x = "x",
  y = "y",
  value = "value",
  type = "raster"
)
```

## Interpretation

The final spatial pattern was not drawn directly. It emerged from repeated local updating. Each grid cell changed according to the model rules, and the system-level structure appeared through iteration.

This illustrates a central lesson of self-organization:

> Global order can arise from local dynamics without a central organizer.

The resulting pattern should not be interpreted as a real chemical or biological structure. It is a teaching model that makes the logic of pattern formation visible.

## Feedback experiment

Feedback controls how strongly local values reinforce or amplify themselves. A low-feedback system may remain relatively smooth. A high-feedback system may develop stronger contrasts or more sharply defined patterns.

```{r feedback-experiment}
low_feedback <- simulate_self_organization(
  grid_size = 30,
  steps = 50,
  diffusion = 0.20,
  feedback = 0.20,
  seed = 2
)

high_feedback <- simulate_self_organization(
  grid_size = 30,
  steps = 50,
  diffusion = 0.20,
  feedback = 0.80,
  seed = 2
)

low_final <- subset(low_feedback, step == max(step))
high_final <- subset(high_feedback, step == max(step))

head(low_final)
head(high_final)
```

## Interpretation of feedback

Increasing feedback changes how strongly the system reinforces local variation. In theoretical terms, feedback can convert small differences into larger patterns.

This is important because many self-organizing systems depend on amplification. Small fluctuations may become meaningful if the system reinforces them. Without feedback, differences may fade. With too much feedback, the system may become unstable or dominated by extreme values.

## Diffusion experiment

Diffusion controls how strongly values spread or smooth across the grid. A low-diffusion system may preserve local variation. A high-diffusion system may become more uniform.

```{r diffusion-experiment}
low_diffusion <- simulate_self_organization(
  grid_size = 30,
  steps = 50,
  diffusion = 0.05,
  feedback = 0.60,
  seed = 2
)

high_diffusion <- simulate_self_organization(
  grid_size = 30,
  steps = 50,
  diffusion = 0.50,
  feedback = 0.60,
  seed = 2
)

low_diff_final <- subset(low_diffusion, step == max(step))
high_diff_final <- subset(high_diffusion, step == max(step))

head(low_diff_final)
head(high_diff_final)
```

## Interpretation of diffusion

Diffusion spreads local influence across space. It can smooth differences, connect nearby regions, and prevent purely isolated behavior.

In many self-organizing systems, diffusion interacts with feedback. Feedback amplifies. Diffusion spreads or smooths. Pattern formation often depends on the balance between the two.

This balance is a key theoretical insight: order does not arise simply because a system has many parts. It arises because the interactions among parts have the right structure.

## Measuring the resulting pattern

Visual inspection is useful, but simple metrics can help compare outputs. The function `measure_emergence()` can summarize diversity, entropy, or temporal change depending on the simulation output.

```{r measure-self-organization}
measure_emergence(
  so,
  value_col = "value",
  time_col = "step"
)
```

Metrics should be interpreted carefully. They do not fully define self-organization, but they provide a way to compare outcomes under different parameter settings.

## Self-organization and life

Self-organization is important in origin-of-life research because early prebiotic systems may have developed structured dynamics before modern genetic control systems existed. Chemical networks, protocell-like compartments, autocatalytic cycles, and metabolic-like processes are often discussed in terms of self-organization [@kauffman1993origins; @maynard1999origins].

This does not mean that self-organization alone explains life. Living systems involve heredity, metabolism, boundary maintenance, adaptation, and evolution. However, self-organization helps explain how structured dynamics may arise before fully developed biological systems.

In this sense, self-organization provides a conceptual bridge between chemistry and biology.

## Self-organization and consciousness

Self-organization is also relevant to consciousness and cognitive science. Neural systems are not centrally controlled in a simple top-down manner. Patterns of activity can arise through distributed interactions among neurons, networks, body, and environment.

This does not mean that self-organization alone explains consciousness. However, it helps frame consciousness as a system-level phenomenon involving coordination, feedback, integration, and dynamic organization.

For this reason, `emergenceModelR` complements projects such as `lifesimulatoR` and `consciousnessModelR`. It provides a general modeling language for thinking about how organized patterns arise in complex systems.

## What the model captures

The function captures several important ideas:

- order can arise without central control;
- local interactions can generate global patterns;
- feedback can amplify small differences;
- diffusion can spread or smooth variation;
- pattern formation depends on parameter balance;
- simulation can reveal structure not obvious from the rules alone.

These features make the model useful for teaching self-organization as a mechanism of emergence.

## What the model does not capture

The model is intentionally simplified. It does not represent:

- real chemical reaction-diffusion systems;
- biological morphogenesis in detail;
- ecological dynamics;
- thermodynamic energy flows;
- real neural self-organization;
- evolutionary adaptation;
- full origin-of-life dynamics.

It is a conceptual model, not a detailed scientific simulation.

## Responsible interpretation

It is better to say:

> The simulation illustrates self-organization-like pattern formation.

than:

> The simulation proves how life or consciousness emerges.

It is better to say:

> The model shows how feedback and diffusion can shape spatial structure.

than:

> The model is a realistic biological system.

Careful language is important because self-organization is sometimes used too broadly. The purpose of the model is to clarify the concept, not to overstate it.

## Educational use

This chapter can support several classroom or self-study questions:

- How does feedback affect pattern formation?
- How does diffusion affect spatial structure?
- What happens when feedback and diffusion are out of balance?
- Can order arise without a central controller?
- How does self-organization differ from random patterning?
- Why is self-organization relevant to life and consciousness?
- What would be needed to make the model more realistic?

These questions help learners understand self-organization as a dynamic process rather than a vague label.

## Key takeaway

Self-organization is a mechanism by which order can arise through internal system dynamics. It does not require a central designer or controller. Instead, patterns form through local interaction, feedback, diffusion, and repeated updating.

`simulate_self_organization()` provides a simplified educational model of this process. It helps learners explore how distributed dynamics can generate structured outcomes, while preserving the distinction between toy simulation and real-world complexity.

## References